\documentclass[12pt,dvipsnames]{amsart}
%prepared in AMSLaTeX, under LaTeX2e
\addtolength{\oddsidemargin}{-.65in} 
\addtolength{\evensidemargin}{-.65in}
\addtolength{\topmargin}{-.4in}
\addtolength{\textwidth}{1.3in}
\addtolength{\textheight}{.8in}

\renewcommand{\baselinestretch}{1.05}

\usepackage{fancyvrb}
\usepackage{palatino}
\usepackage[final]{graphicx}
\usepackage{amssymb,enumitem,xspace,xcolor}
\usepackage{hyperref}
\hypersetup{pdfauthor={Ed Bueler},
            pdfcreator={pdflatex},
            colorlinks=true,
            citecolor=blue,
            linkcolor=red,
            urlcolor=blue,
            }

\newtheorem*{thm}{Theorem}
\newtheorem*{defn}{Definition}
\newtheorem*{example}{Example}
\newtheorem*{problem}{Problem}
\newtheorem*{remark}{Remark}

\newcommand{\br}{\mathbf{r}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}

\newcommand{\eps}{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\lam}{\lambda}
\newcommand{\lap}{\triangle}

\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}

\newcommand{\image}{\operatorname{im}}
\newcommand{\onull}{\operatorname{null}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\range}{\operatorname{range}}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\prob}[1]{\bigskip\noindent\textbf{#1.}\quad }
\newcommand{\exer}[2]{\prob{Exercise #2 in Lecture #1}}

\newcommand{\pts}[1]{(\emph{#1 pts}) }
\newcommand{\epart}[1]{\medskip\noindent\textbf{(#1)}\quad }
\newcommand{\ppart}[1]{\,\textbf{(#1)}\quad }

\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\Python}{\textsc{Python}\xspace}
\newcommand{\Julia}{\textsc{Julia}\xspace}

\newcommand{\ds}{\displaystyle}

\DefineVerbatimEnvironment{mVerb}{Verbatim}{numbersep=2mm,
frame=lines,framerule=0.1mm,framesep=2mm,xleftmargin=4mm,fontsize=\footnotesize}


\begin{document}
\scriptsize \noindent Math 614 Numerical Linear Algebra (Bueler) \hfill \emph{assigned 13 October 2025}
\normalsize\medskip

\Large\centerline{\textbf{Assignment \#6}}
\large
\medskip

\centerline{\textbf{Due Wednesday 22 October, at the start of class}}
\medskip
\normalsize

\thispagestyle{empty}

\bigskip

\noindent Please read Lectures 10, 11, 12, and 13 in the textbook \emph{Numerical Linear Algebra}, SIAM Press 1997, by Trefethen and Bau.

\bigskip
\noindent \textsc{Do the following exercises from the textbook}:

\begin{itemize}[itemsep=4pt]
\item \textbf{Exercise 11.3}
\item \textbf{Exercise 12.2}
\item \textbf{Exercise 13.4}  \quad\, \begin{minipage}[t]{0.68\textwidth}   \emph{Any computer you touch has $\eps_{\text{machine}}\approx 10^{-16}$, so do part (a) with floating point numbers as usual.  I was able to get Matlab to do part (b) by starting with} {\small \texttt{x = sym('0')}} \emph{and then doing Newton's method the usual way at the command line.  For instance, if I were approximating $\sqrt{2}$ by Newton's method then I might start with} \, {\small \texttt{x = sym('1')}} \emph{and then repeat} \, {\small \texttt{x = x - (x$\wedge$2-2) / (2*x)}}.  \emph{What is Trefethen's point in showing you this calculation?}\end{minipage}
\end{itemize}


\medskip
\noindent \textsc{Do the following additional problems.}

\medskip
\noindent \emph{The Matlab built-in} \texttt{qr()} \emph{computes the QR factorization using Householder reflectors (Lecture 10).  Generally you should use it when QR is needed, but of course you can check it against} \texttt{mgs()}, \emph{which you wrote on Assignment \#5 from Lecture 8.}

\prob{P14}  By applying Matlab's backslash command, reproduce Figure 11.1.  Then, by applying Algorithm 11.2, using the \texttt{qr} and backslash commands, reproduce Figure 11.2.  Please make at least a modest effort to capture the appearance of these Figures.  (\emph{Hints.}  Note \,\texttt{axis off}\, creates a clean picture without ticks and axes labels.  You can put back the black axes using\, \texttt{plot([-6 6],[0 0],'k')} and similar.)

\prob{P15}  Suppose $A$ is a $101\times 101$ matrix with $\|A\|_2=50$ and $\|A\|_F=51$.  Give the sharpest possible lower bound on the 2-norm condition number of $A$.  (\emph{Hint.  Write everything in terms of singular values, and then think about best cases for $\kappa_2(A)$.})



\end{document}
