\documentclass[12pt]{amsart}
\addtolength{\topmargin}{-0.6in} % usually -0.25in
\addtolength{\textheight}{1.1in} % usually 1.25in
\addtolength{\oddsidemargin}{-0.7in}
\addtolength{\evensidemargin}{-0.7in}
\addtolength{\textwidth}{1.5in} %\setlength{\parindent}{0pt}

\newcommand{\normalspacing}{\renewcommand{\baselinestretch}{1.04}\tiny\normalsize}

\normalspacing

% macros
\usepackage{amssymb,xspace}
\usepackage{tikz}
\usepackage[pdftex,colorlinks=true]{hyperref}


\newtheorem*{thm}{Theorem}
\newtheorem*{lem}{Lemma}

\newcommand{\mtt}{\texttt}
\newcommand{\mtl}[1]{{\texttt{>>#1}}}
\usepackage{alltt}
\usepackage{fancyvrb}

\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}

\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\RR}{{\mathbb{R}}}
\newcommand{\ZZ}{{\mathbb{Z}}}
\newcommand{\ZZn}{{\mathbb{Z}}_n}
\newcommand{\NN}{{\mathbb{N}}}

\newcommand{\eps}{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\lam}{\lambda}
\newcommand{\ip}[2]{\mathrm{\left<#1,#2\right>}}
\newcommand{\erf}{\operatorname{erf}}

\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}
\newcommand{\Arg}{\operatorname{Arg}}

\newcommand{\Span}{\operatorname{span}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\range}{\operatorname{range}}
\newcommand{\trace}{\operatorname{tr}}
\newcommand{\Null}{\operatorname{null}}

\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\pylab}{\textsc{pylab}\xspace}
\newcommand{\longMOP}{\textsc{Matlab}\big|\textsc{Octave}\big|\textsc{pylab}\xspace}
\newcommand{\MOP}{\textsc{M}\big|\textsc{O}\big|\textsc{p}\xspace}

\newcommand{\prob}[1]{\bigskip\noindent\large\textbf{#1.} \normalsize}
\newcommand{\bookprob}[1]{\bigskip\noindent\large\textbf{Exercise #1.} \normalsize}
\newcommand{\probpart}[1]{\smallskip\noindent\textbf{(#1)}\quad }
\newcommand{\aprobpart}[1]{\textbf{(#1)}\quad }


\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}


\begin{document}
\scriptsize \noindent Math 614 Numerical Linear Algebra (Bueler)
\thispagestyle{empty}

\bigskip
\Large\textbf{\centerline{Final Exam: three algorithms}}

\medskip
\large\textbf{\centerline{Wednesday, 13 December, 1:00pm--3:00pm, Chapman 107}}

\normalsize
\bigskip
The in-class Final Exam will be a bit different, but short.  I would like you to summarize three major algorithms which we have seen during the semester, chosen from the options below.

\smallskip
You will have this document in your hand when you do the final, but otherwise you must remember what you want to say.  \textbf{You may NOT bring your own notes, or the textbook, to the final.}  However, you are \textbf{strongly encouraged to draft and practice your planned summaries.}  Make sure to read relevant sections of the textbook for your topics.  Feel free to get feedback on your drafts from other students or faculty/friends/family/pets, or me.  Then think through, in preparation, how you will remember enough detail so as to recreate the summary during the Final Exam itself.

\medskip
\centerline{\textbf{Please choose one of categories A,B,C from the lists below.}}

\centerline{\textbf{For each one, write a Summary on a separate sheet.}}

\smallskip
Each summary should be at most a page.  FIXME Equations are important, plus a few (at most three) sketches as needed, but please write complete sentences for the major ideas.  A Summary which is \emph{only} equations or sketches cannot achieve a good grade.

\smallskip
FIXME Except for special cases, a Summary should \textbf{state the problem} (DE and initial/boundary conditions), comment on or supply an \textbf{exact solution}, state one or two \textbf{FD discretization}(s), summarize the \textbf{truncation error}, and summarize \textbf{stability}.  Then it should consider and fill-in any relevant additional ideas for the problem and the chosen method; see ``Consider:'' below for the specific problems.  The goal is \emph{not} to prove anything, but to accurately and precisely summarize what we know---or we easily could know---about the problem and the available method(s), especially in advance of writing a code.

\smallskip
For each Summary your goal is to write something like a neat 10 or 15 minute white-board presentation of how you would solve the problem numerically, as you might give to a hypothetical fellow graduate student who has not seen that particular problem but who has taken similar-level classes.  Give an executive summary, and remember to point out the basic things!

\medskip

\newcommand{\ecomment}[1]{ \hfill \mbox{\emph{#1}} }

\renewcommand{\labelenumi}{ODE \circled{\arabic{enumi}}}
\begin{enumerate}
\setlength{\itemsep}{4pt}
\item A general linear 2nd-order ODE BVP. \ecomment{eqn (2.64)}

\noindent Consider: how to solve the linear system.
\item The nonlinear 2nd-order ODE BVP for a pendulum.  \ecomment{eqn (2.77)}

\noindent Consider: Newton's method, how to solve the linear Newton steps.
\item A non-homogeneous, constant-coefficient linear ODE IVP system.  \ecomment{eqn (5.6)}

\noindent Consider: matrix exponentials, compare $\ge 2$ schemes, absolute stability regions.
\end{enumerate}

\renewcommand{\labelenumi}{PDE \circled{\alph{enumi}}}
\begin{enumerate}
\setlength{\itemsep}{4pt}
\item The Poisson problem PDE in 2D. \ecomment{eqn (3.5)}

\noindent Consider: ordering the unknowns, how to solve the linear system.
\item The time-dependent heat equation PDE in 1D. \ecomment{eqn (9.1)}

\noindent Consider: compare $\ge 2$ schemes.
\item The time-dependent scalar advection equation PDE in 1D.  \ecomment{eqn (10.1)}

\noindent Consider: compare $\ge 2$ schemes.
\end{enumerate}

\end{document}

