\documentclass[12pt]{amsart}
%prepared in AMSLaTeX, under LaTeX2e
\addtolength{\oddsidemargin}{-.65in} 
\addtolength{\evensidemargin}{-.65in}
\addtolength{\topmargin}{-.4in}
\addtolength{\textwidth}{1.3in}
\addtolength{\textheight}{.75in}

\renewcommand{\baselinestretch}{1.05}

\usepackage{verbatim} % for "comment" environment

\usepackage{palatino}

\newtheorem*{thm}{Theorem}
\newtheorem*{defn}{Definition}
\newtheorem*{example}{Example}
\newtheorem*{problem}{Problem}
\newtheorem*{remark}{Remark}

\usepackage{fancyvrb,xspace}

\usepackage[final]{graphicx}

% macros
\usepackage{amssymb}

\usepackage{hyperref}
\hypersetup{pdfauthor={Ed Bueler},
            pdfcreator={pdflatex},
            colorlinks=true,
            citecolor=blue,
            linkcolor=red,
            urlcolor=blue,
            }

\newcommand{\br}{\mathbf{r}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}

\newcommand{\eps}{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\lam}{\lambda}
\newcommand{\lap}{\triangle}

\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}

\newcommand{\image}{\operatorname{im}}
\newcommand{\onull}{\operatorname{null}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\range}{\operatorname{range}}
\newcommand{\trace}{\operatorname{tr}}

\newcommand{\prob}[1]{\bigskip\noindent\textbf{#1.}\quad }
\newcommand{\exer}[2]{\prob{Exercise #2 in Lecture #1}}

\newcommand{\pts}[1]{(\emph{#1 pts}) }
\newcommand{\epart}[1]{\medskip\noindent\textbf{(#1)}\quad }
\newcommand{\ppart}[1]{\,\textbf{(#1)}\quad }

\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\Python}{\textsc{Python}\xspace}
\newcommand{\Julia}{\textsc{Julia}\xspace}

\newcommand{\ds}{\displaystyle}

\DefineVerbatimEnvironment{mVerb}{Verbatim}{numbersep=2mm,
frame=lines,framerule=0.1mm,framesep=2mm,xleftmargin=4mm,fontsize=\footnotesize}



\begin{document}
\scriptsize \noindent Math 614 Numerical Linear Algebra (Bueler) \hfill \emph{assigned 9 October 2023}
\normalsize\medskip

\Large\centerline{\textbf{Assignment 6}}
\large
\medskip

\centerline{\textbf{Due Friday 20 October 2023, at the start of class}}
\medskip
\normalsize

\thispagestyle{empty}

\bigskip
\noindent Please read Lectures 8,9,10,11,12 in the textbook \emph{Numerical Linear Algebra} by Trefethen and Bau.  This Assignment mostly covers Gram-Schmidt QR, Householder reflectors and QR, and least squares.

\bigskip
\noindent \textsc{Do the following exercises} from Lecture 8:

\begin{itemize}
\item \textbf{Exercise 8.3}
\end{itemize}

\bigskip
\noindent \textsc{Do the following exercises} from Lecture 9:

\begin{itemize}
\item \textbf{Exercise 9.3}
\end{itemize}

\bigskip
\noindent \textsc{Do the following exercises} from Lecture 10:

\begin{itemize}
\item \textbf{Exercise 10.2}
\item \textbf{Exercise 10.3}
\end{itemize}


\bigskip
\noindent \textsc{Do the following additional exercises.}

\medskip
\noindent \emph{The Matlab built-in} \texttt{qr()} \emph{computes the QR factorization using Householder reflectors (Lecture 10).  In the next two problems, go ahead and use it.}

\prob{P11}  By applying Matlab's backslash command, reproduce Figure 11.1.  By applying Algorithm 11.2, using the \texttt{qr} and backslash commands, reproduce Figure 11.2.  Please make at least a modest effort to duplicate the appearance of these Figures.  (\emph{Hints.}  Note \,\texttt{axis off}\, creates a clean picture without ticks and axes labels.  Then you can put back the axes themselves using \texttt{plot([-6 6],[0 0],'k')} and similar.)


\prob{P12}  \emph{While we have used QR to solve linear systems, here we see that QR factorization has a completely different application.  For more, see Lectures 24--29.}

\epart{a} By googling for ``unsolvable quintic polynomials'' or similar, confirm that there is a theorem which shows that fifth and higher-degree polynomials cannot be solved using finitely-many operations (including roots, a.k.a.~``radicals'').  In other words, there is no finite formula for the solutions (``roots'') of such polynomial equations.  Who proved this theorem?  When?  Show a quintic polynomial for which it is known that there is no finite formula.  (\emph{You do \emph{not} need to prove your claim!})

\clearpage\newpage
\epart{b} At the Matlab command line, try the following:
\begin{mVerb}
>> A = randn(5,5);  A = A' * A;    % create a random 5x5 symmetric matrix
>> A0 = A;                         % save a copy of the original A
>> [Q, R] = qr(A);  A = R * Q
...                                % repeat about 10 times
>> [Q, R] = qr(A);  A = R * Q
\end{mVerb}
We start with a random, symmetric $5\times 5$ matrix $A_0$ and then generate a sequence of new matrices $A_i$.  Specifically, each matrix $A_i$ is factored
    $$A_i = Q_i R_i$$
and then the next matrix $A_{i+1}$ is generated by multiplying-back in reversed order:
    $$A_{i+1} = R_i Q_i.$$
What happens to the matrix entries when you iterate at least 10 times?  (\emph{Perhaps also use a} \texttt{for} \emph{loop to see a strong effect from e.g.~100 iterations.})  What do you observe about this sequence of $A_i$?  Now compare \verb|sort(diag(A))| to \verb|sort(eig(A0))|.

\epart{c} To see a bit more of what is going on in part \textbf{(b)}, show that
    $$A_{i+1} = Q_i^* A_i Q_i.$$
This shows $A_{i+1}$ has exactly the same eigenvalues as $A_i$\,; \,explain why.

\epart{d} Write a few sentences which relate part \textbf{(a)} to what happens in parts \textbf{(b)} and \textbf{(c)}.  (\emph{Hint.  Try to relate the two parts by yourself first.  Then read ``A Fundamental Difficulty'' in Lecture 25 to confirm your understanding.})


\end{document}
